Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-27 paper.tex
tunnel characteristics, not finished yet

\subsection{Main idea \red{VV dodelam}}

The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand in the tunnel.
As the traversability may depends on the initial position of the ligand, especially its rotation, it is necessary to compute
the trajectories from multiple starting configurations $\qinit$.
The starting configuration $\qinit$ is generated by randomly sampling the vicinity of the first sphere $c_1$ of the tunnel.
A sample $q$ is created by generating its translation part randomly around $t_1$ in the distance $R$.
The rotation part of $q$ and conformation index are set randomly, the scale is set to $\smin$.
If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.

Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the next section.

@@

%\section{Ligand traversability}

The traversability study proposed in this paper is achieved by computing many trajectories for the given ligand and tunnel, which
is solved using a novel RRT-based planner.
The main principle of the RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand
conformations.
The main loop of the method is described in Alg.~\ref{alg::main}.
%The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
%We propose a RRT-based method to compute the trajectories of the ligand.
%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
%suffer from the well known narrow passage problem~\cite{hannaWIS}.
%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
%Instead, they are generated using the spheres of the tunnel.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a tunnel sphere.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
around the sphere $t_v \in T$.
%around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

\subsection{Computing trajectories}

The result of the planner is the tree $\T$ of collision-free configurations.
The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the neearest node in the tree towards $\qgoal$, is found.
The path $P=(q_i), q_i \in \C$ is represented by a sequence of the collision-free configurations.

Due to stochastic nature of RRT, the computed trajectories may differ in each run.
Moreover, they are influenced by the initial configuration $\qinit$, especialy by the rotation of the ligand.
To evaluate traversability of the given tunnel, multiple starting configurations are generated and for each
of these configurations, multiple trajectories are computed.
This results in a set of $n$ trajectories $P_i, i=1,\ldots,n$.
These trajectories are then used to evaluate tunnel characteristics, which is described in the next section.

@@

\red{VV is significantly rewriting this section}

There basic characteristics of the tunnel $T$ are computed from the trajectories: the accessiblity, throughput and the scaling factor.
These characteristics are computed for each sphere $(c_i,r_i) \in T$ of the tunnel.
This requires to determine which parts of the tunnel were accessed by the trajectories.

>>>>

\subsection{Main idea}

The analysis of tunnel traversability is based on computation of multiple trajectories of the ligand in the tunnel.
As the traversability may depends on the initial position of the ligand, especially its rotation, it is necessary to compute
the trajectories from multiple starting configurations $\qinit$.
The starting configuration $\qinit$ is generated by randomly sampling the vicinity of the first sphere $c_1$ of the tunnel.
A sample $q$ is created by generating its translation part randomly around $t_1$ in the distance $R$.
The rotation part of $q$ and conformation index are set randomly, the scale is set to $\smin$.
If the sample $q$ is collision-free, it can be considered as a new starting configuration $\qinit$.
Then, multiple trajectories are computed starting from $\qinit$ using the method proposed in the next section.

@@

The trajectories are computed using an extension of the RRT~\cite{lavalleRRT} planner.
The RRT planner is extended by guided sampling, scaling of the ligands atoms and using predefined set of ligand conformations.
The main loop of the method is described in Alg.~\ref{alg::main}.
%The spheres representing the tunnel are used to generate the random samples for the proposed RRT-based planner.
%We propose a RRT-based method to compute the trajectories of the ligand.
%As the tunnels are usually computed on proteins without ligands, their are usually very narrow and the classic RRT would
%suffer from the well known narrow passage problem~\cite{hannaWIS}.
%To cope with the narrow passage problem, the random samples are not generating from the whole configuration space.
%Instead, they are generated using the spheres of the tunnel.
In each iteration, a random sample $\qrand$ is generated, its nearest node $\qnear\in\T$ in the tree is found.
The nearest-neighbor search between the $\qrand$ and the tree is realized using the weighted 6D Euclidean metric considering
both 3D rotation and 3D translation.

To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of sphere of the tunnel.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
around the sphere $t_v \in T$.
After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

@@

The result of the planner is the tree $\T$ of collision-free configurations.
The path between $\qinit$ (root of the tree) and $q'$, where $q'$ is the neearest node in the tree towards $\qgoal$, is found.
The path $P=(q_i), q_i \in \C$ is represented as a sequence of the collision-free configurations.
Let $d(P,i)$ denote the 3D Euclidean distance

The path $P$ reaches a particular sphere $t_i$ in the tunnel if the 3D Euclidean distance between the sphere and its nearest configuration
in the path is less than the threshold $T$.
TODO VV dodelam

@@

The traversability of the tunnel $T$ is evaluated using all computed trajectories $P_i$. 
Three basic characteristics of the tunnel are computed from the trajectories: the accessiblity, throughput and the scaling factor.
These characteristics are computed for each sphere $(c_i,r_i) \in T$ of the tunnel.
This requires to determine which parts of the tunnel were accessed by the trajectories.
